Breakpoint Models Show No Evidence of Recreational Adaptation to Increasing Wildfire Smoke in the American West

### 2021-05-22 MarkdownDocuments/Manuscript_Real.Rmd
updated plots in code and manuscript

![image alt text](../Figures/smkplt.png)

@@

### Visitation Data
Visitation data was obtained through the [National Park Service Visitor Use Statistics Portal](https://irma.nps.gov/STATS/). These data are generally collected via car counters, permit information, and concessionaire reporting. Methods of collection vary from park to park. Unit specific information can be found on the National Park Service Visitor Use Statistics Portal. Although collection methods are inconsistent across the sample, all data was taken as reported. We retrieved monthly data for all national park units in the Intermountain and Pacific West regions for years 1980 through 2019. To reduce issues regarding park seasonality, we then subset our data to only the three month high season associated with each park. For example, only Summer (June, July, Aug) data was used for Yellowstone National Park.

@@

We formulated this baseline autoregressive model using a bayesian heierarchical framework. This approach is appropriate here as it allows us to account between park variation in visitation trends while acknowledging the interconnectedness of these trends through partial pooling [@statrethinkingbook]. For each park ($j$) we estimate a parameter value ($\bm{\nu1...k}$) representing the trend in visitation from one, two, and three years prior ($t-1,2,...k$), to the month we are estimating ($i$). Finally, we estimate both global ($\alpha0$) and park specific ($\bm{\beta0}_{j}$) intercepts. We modeled these data using a negative binomial distribution to allow us to estimate an overdispersion paremeter ($\phi$) rather than assuming the dispersion is equal to the mean.

@@

$$\bm{\beta0}_{j} \sim Norm(0,\sigma_{j})$$

teeeext

![image alt text](../Figures/bpplot.png)

@@

$$\bm{\beta0}_{j} \sim Norm(0,\sigma_{j})$$

### Validation
Text about model checking

# Results

>>>>

![image alt text. Righ-hand y axis shows the standardized values of smoke ](../Figures/smkplt.png)

@@

### Visitation Data
Visitation data were obtained on the monthly scale through the [National Park Service Visitor Use Statistics Portal](https://irma.nps.gov/STATS/). These data are generally collected via car counters, permit information, and concessionaire reporting. Methods of collection vary from park to park. Unit specific information can be found on the National Park Service Visitor Use Statistics Portal. Although collection methods are inconsistent across the sample, all data was taken as reported. We retrieved monthly data for all national park units in the Intermountain and Pacific West regions for years 1980 through 2019. To reduce issues regarding park seasonality, we then subset our data to only the three month high season associated with each park. For example, only Summer (June, July, Aug) data was used for Yellowstone National Park.

@@

We formulated this baseline autoregressive model using a Bayesian hierarchical framework. This approach is appropriate here as it allows us to account between park variation in visitation trends while acknowledging the interconnectedness of these trends through partial pooling [@statrethinkingbook]. For each park ($j$) we estimate a parameter value ($\bm{\nu1...k}$) representing the trend in visitation from one, two, and three years prior ($t-1,2,...k$), to the month we are estimating ($i$). Finally, we estimate both global ($\alpha0$) and park specific ($\bm{\beta0}_{j}$) intercepts to yield equation 1 below. We modeled these data using a negative binomial distribution to allow us to estimate an overdispersion parameter ($\phi$) rather than assuming the dispersion is equal to the mean.

@@

$$\bm{\beta0}_{j} \sim Norm(0,\sigma)$$

After validating that the model above captures the baseline trends in national park visitation from 1980-2019, we built upon it to estimate the if a critical threshold for visitation response to wildfire smoke exists in these data and if so, what the impact of wildfire smoke is on visitation post-threshold. To do this, we developed a series of three breakpoint models, allowing for three increasingly extreme breakpoints (i.e. critical thresholds). We standardized these smoke data, and thus the thresholds as well, by dividing by two standard deviations following the recommendation made by Gelman [-@gelman2008scaling]. A threshold of 1.0 can then be interpreted as two standard deviations above the mean value of smoke in the data. A visual comparison of the observed actual smoke values and standardized values can be seen in Fig. 1.

As shown conceptually in Fig. 2, we estimated the effect of smoke on visitation before ($\bm{\beta1}$) and after ($\bm{\beta2}$) each breakpoint (BP). Just as with the autoregressive terms, we estimated unique paramater values for each park ($j$) in a heierarchical framework. The complete model used for each of the three breakpoint values is then as seen in equation 2.

![image alt text](../Figures/bpplot2.png)

@@

$$\bm{\beta0}_{j} \sim Norm(0,\sigma)$$

Following the recommendation given by Gelman et al. [-@gelman2008weakly] for producing stable, conservative estimates, we specified long-tailed regularizing priors for all autoregressive terms, the global intercept value, and the overdispersion parameter. We used normally distributed regularizing priors for estimating the variance parameter ($\sigma$) and the pre and post-breakpoint smoke effects to allow for more efficient sampling while generative within-sample predictions at high smoke values.

### Validation
Due to the considerable computational time (about 1 day) needed to run the program described above, we validated the results of both the autoregressive-only and breakpoint models using within-sample posterior predictive checks. All models were first inspected to ensure adequate mixing of Markov chains and $\hat{R}$ values below 1.05 for all parameter estimates.

![image alt text](../Figures/ARVal.png)

# Results
While some between-park variation exists, our overall study shows no evidence for recreational visitor response to wildfire smoke in national parks in the American West.
